---
title: "phyloseq_analysis"
output: html_document
---

# Setup
## Get libraries

```{r, message=FALSE}
library(dplyr)
library(dada2)
library(phyloseq)
library(Biostrings)
library(ggplot2)
```

## Compile from metadata.txt (editable), and all other files (not editable)
```{r}
base = "."
data_dir = file.path(base, "data")
out_dir = file.path(base, "results")
ids.file = "sample_ids_JoelyReyes.txt"

metadata.file = file.path(out_dir, "metadata.txt")
ASVfasta.file = file.path(out_dir, "ITS.fa")
ASVtable.file = file.path(out_dir, "ITS_ASVtable.txt")
taxatable.file = file.path(out_dir, "ITS_taxonomy.txt")

metadata = read.table(metadata.file, sep = "\t", check.names = F, header = T)
rownames(metadata) = metadata[, "sample"]
metadata = metadata[,1:ncol(metadata)]

ASVtable = read.table(ASVtable.file, sep = "\t", check.names = F, header = T)
rownames(ASVtable) = ASVtable[, "asv_seq"]
ASVtable = as.matrix(ASVtable[,2:ncol(ASVtable)])

taxatable = as.matrix(read.table(taxatable.file, sep = "\t", check.names = F, header = T))
rownames(taxatable) = taxatable[, "asv_seq"]
taxatable = taxatable[,2:ncol(taxatable)]

refseq = readDNAStringSet(ASVfasta.file, format="fasta",
                          nrec = -1L, skip = 0L, seek.first.rec = FALSE,
                          use.names = TRUE, with.qualities = FALSE)
phyloseq <- phyloseq(otu_table = otu_table(ASVtable, taxa_are_rows = TRUE),
                     sample_data = sample_data(metadata),
                     tax_table = tax_table(taxatable))
save(phyloseq, file = file.path(out_dir, "ITS.phyloseq.RData"))
phyloseq
```

# Alpha Diversity plots
## Prune data
```{r}
phyloseq = prune_taxa(taxa_sums(phyloseq) > 0, phyloseq)
```

## Default
```{r}
plot_richness(phyloseq)
```

For now, lets take a look at Chao1 and Shannon diversity indices

```{r}
plot_richness(phyloseq, measures=c("Chao1", "Shannon"))
```

## Soil-Treatment combination
```{r}
plot_richness(phyloseq, x="soil_treatment", measures=c("Chao1", "Shannon"))
```

## Time Point
```{r}
plot_richness(phyloseq, x="time", measures=c("Chao1", "Shannon"))
```

Graph was colored by soil-treatment combination. However, this showed no significant difference in alpha diversity between the soil-treatment combinations. Do note there were less T0 soil samples, so there may not be as much power to detect differences.

## Soil and Treatment
```{r}
plot_richness(phyloseq, x="treatment", measures=c("Chao1", "Shannon"), color="soil")
```

Also made a graph to compare the alpha diversity between the soil types, colored by the treatment:

```{r}
plot_richness(phyloseq, x="soil", measures=c("Chao1", "Shannon"), color="treatment")
```


## Soil-Treatment combination and Time Point
```{r}
plot_richness(phyloseq, x="soil_treatment", measures=c("Chao1", "Shannon"), color="time")
```

# Ordination
## Default NMDS with Bray distance
```{r}
phyloseq.ord <- ordinate(phyloseq, method="NMDS", distance="bray")
p1 = plot_ordination(phyloseq, phyloseq.ord, type="taxa", color="Phylum", title="taxa")
print(p1)
```

## Only with the top 5 phyla
```{r}
phylum.sum = tapply(taxa_sums(phyloseq), tax_table(phyloseq)[, "Phylum"], sum, na.rm=TRUE)
top5phyla = names(sort(phylum.sum, TRUE))[1:5]
phyloseq = prune_taxa((tax_table(phyloseq)[, "Phylum"] %in% top5phyla), phyloseq)
phyloseq.ord = ordinate(phyloseq, method="NMDS", distance="bray")
p2 = plot_ordination(phyloseq, phyloseq.ord, type="taxa", color="Phylum", title="taxa")
print(p2)
```

## Plot with facet wrap
```{r}
p3 = plot_ordination(phyloseq, phyloseq.ord, type="taxa", color="Phylum", title="taxa") + facet_wrap(~Phylum, 3)
print(p3)
```

## Plot with shapes
```{r}
p4 = plot_ordination(phyloseq, phyloseq.ord, type="samples", color="soil_treatment", shape="time")
p4 + geom_polygon(aes(fill=soil_treatment)) + geom_point(size=5) + ggtitle("samples")
```